---
title: "Logbook parkinson dysphonia"
author: "L T Stein"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The data is extracted from Kaggle and its article is published on ncbi by Max A. Little, Patrick E. McSharry, Eric J. Hunter, Lorraine O. Ramig (2008). 'Suitability of dysphonia measurements for telemonitoring of Parkinson's disease', *IEEE Transactions on Biomedical Engineering (to appear)*.


---

# Read data


First a look of what the data looks like from a .csv file. The data is divided
by header and is comma separated. 


```{r}
patient.data <- read.table("parkinsons.data", header = T, sep = ",")
head(patient.data)
```

Verify if everything is read correctly.

```{r}
cat(ncol(patient.data), "columns and", nrow(patient.data), "rows")
```

There are 2 nominal and 22 numeric columns.


Showing the amount of recordings of healthy people and sick people. The original labels
of one and zero are replaced respectfully with sick and healthy 

```{r}
# Adding labels instead of "1" and "0"
patient.data$status[patient.data$status == "1"] <- "sick"
patient.data$status[patient.data$status == "0"] <- "healthy"


amount.sick.recordings <- sum(patient.data$status == "sick")
amount.healthy.recordings <- sum(patient.data$status == "healthy")
cat(" There are [", amount.sick.recordings, "] recordings of Parkinson patients.\n",
    "There are [", amount.healthy.recordings, "] recordings of healthy people.\n",
    "In total there are: [", amount.sick.recordings + amount.healthy.recordings, "] recordings.")
```


In order to make it more clear what the data means, the columns are put into a table.


```{r}
columns <- colnames(patient.data)
definitions <- read.csv("definitions.csv", header = FALSE, sep = ",")
attributes <- data.frame(Attributes = c(columns),
                          Unit = c("-", rep("Hz", 3), "%", "ms", rep("-", 4), "dB", rep("-", 13)),
                         Datatype = c(sapply(patient.data, class)[1]),
                         Definition = c(definitions[1]))
attributes
```

**Retrospective**
All the data is read correctly with the corresponding amount of rows and columns
as the article suggests. 


---

## Explorative Data Analysis

It's important to check if there are any NA or missing values.

```{r}
cat("There are", sum(is.na(patient.data)), "missing values")
```

**Retrospective**
No NA values are found.


By applying a summary to the dataset's dataframe, its statistical meanings are returned.

```{r}
summary(patient.data)
```


# Bibliography

Link about EDA
https://r4ds.had.co.nz/exploratory-data-analysis.html 
Link article
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5434464/ 

